---
title: "March Madness"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")
```


```{r}
library(hoopR)
library(elo)
library(gtools) 
library(ggplot2) # Load ggplot2
library(ggExtra) # Load ggExtra
library(dplyr) # Load dplyr
library(ggalluvial) # Load ggalluvial
library(tidyverse) # Load tidyverse
library(ggimage) # Load ggimage
library(GGally) # Load GGally
library(ggrepel) # Load ggrepel
source("~/Desktop/Mod 3/Mod3 Sport Analytics/Week 2/simple_dark_theme-2.R")
```

```{r}
ls("package:hoopR") |> grep("espn_mbb_", x = _, value = TRUE)
```


```{r}
tictoc::tic()
progressr::with_progress({
  mbb_pbp <-  hoopR::load_mbb_pbp()
})
tictoc::toc()
```

```{r}
summary(mbb_pbp)
```

```{r}
mbb_schedule_25 <- load_mbb_schedule(2025)
team_box_2k25 <- load_mbb_team_box(2025)
player_box_2k25 <- load_mbb_player_box(2025)
rankings <- espn_mbb_rankings()
standings_2k25 <- espn_mbb_standings(2025)

```


Game Results 

* Home Win - 1
* Away Win - 0 
* Overtime - 0.5 - Need to figure out how to add this in (Met with Martin on 2/10)

```{r}
mbb_2025 <- load_mbb_pbp(2025)
clean_mbb_25 <- mbb_2025 %>%
  filter(!is.na(home_score)) %>%
  mutate(home_win = ifelse(home_score > away_score, 1, 0))
```
```{r}
summary(mbb_2025$period_number)
```

```{r}
save(mbb_pbp, player_box_2k25, rankings, standings_2k25, team_box_2k25, mbb_schedule_25, mbb_2025, file = "mbb_data.rda")
```

Flip data frame on itself to get most recent game instead of the first game to get more accurate ELO

```{r}
temp_1 <- clean_mbb_25[nrow(clean_mbb_25):1,]

games_1 <- temp_1[!duplicated(temp_1$game_id), ]
```

```{r}
ot_games <- unique(clean_mbb_25[which(clean_mbb_25$period_number > 2), c("game_id")])$game_id
```

```{r}
games_1$home_win[which(games_1$game_id %in% ot_games)] <- 0.5
```


```{r}
elo_2025 <- elo.run(
  home_win ~ home_team_name + away_team_name, 
  data = games_1, 
  k = 30
)

final_2025_elos <- final.elos(elo_2025)
```

```{r}
priors_2026 <- (final_2025_elos * 0.75) + (1500 * 0.25) #New Season; different squads/lineups; I want to try to average out everybody just a little bit

view(priors_2026)
```



```{r}
mbb_2026 <- load_mbb_pbp(2026)
clean_mbb_26 <- mbb_2026 %>% 
  filter(!is.na(home_score)) %>%
  mutate(home_win = ifelse(home_score > away_score, 1,0))

ot_games <- unique(clean_mbb_26[which(clean_mbb_26$period_number > 2), c("game_id")])$game_id

temp_2 <- clean_mbb_26[nrow(clean_mbb_26):1,]

games_2 <- temp_2[!duplicated(temp_2$game_id), ]

games_2$home_win[which(games_2$game_id %in% ot_games)] <- 0.5


```

```{r}
elo_2026 <- elo.run(
  home_win ~ home_team_name + away_team_name, 
  data = games_2, 
  k = 30,
  priors = priors_2026 
)
```

```{r}
summary(mbb_2026$period_number)
```

```{r}
df_elo_26 <- as.data.frame(final.elos(elo_2026))
df_elo_25 <- as.data.frame(final.elos(elo_2025))
```


Keep for later to look at teams for graphing

```{r}
teams_26 <- espn_mbb_teams(2026) 
head(teams_26)
```
get unique column value for display names, upload it to a CSV and then use AI to populate with conferences, then load back in. 

```{r}
unique_teams <- unique(teams_26$display_name)
unique_teams
```

```{r}
library(readr)
write_csv(teams_26, "teams_2026.csv")

conferences_26 <- read_csv('~/Desktop/Mod 3/Mod3 Sport Analytics/Sports Analytics Proj./teams_2026_final.csv')
conferences_26
```

```{r}
load_mbb_schedule(2026)
schedule_26 <- load_mbb_schedule(2026)
head(schedule_26)
```

```{r}
conferences_26_clean <- conferences_26 %>% select(-c('display_name','short_name', 'nickname', 'logos_href_3','logos_href_4', 'logos_href_5', 'logos_href_6', 'logos_href_7', 'logos_href_8', 'logos_href_9', 'logos_href_10', 'logos_href_11', 'logos_href_12', 'logos_href_13', 'logos_href_14'))
```

```{r}
save(espn_mbb_pbp, temp_1, temp_2, games_1, games_2, mbb_2026, mbb_2025, df_elo_25, df_elo_26, conferences_26_clean, schedule_26,   file = "mbb_data_2.rda")
```

```{r}

df_elo_26$team_name <- rownames(df_elo_26)

colnames(df_elo_26)[1] <- "elo_rating"

merged_mbb_clean <- merge(
    x = df_elo_26,
    y = conferences_26_clean,
    by.x = 'team_name',      
    by.y = 'team',   
    all.x = FALSE,           
    all.y = TRUE             
)


```

Graph out all of the teams by their conference and their ELOs on the season to this point. 

```{r}
g_1 <- ggplot(merged_mbb_clean, 
               aes(x = elo_rating, 
                   y = conference_2025_26  
                   )) + 
  geom_point(alpha = 0.3) + 
 geom_image(aes(image = logo_dark), size = 0.05, asp = 16 / 9) + 
  dark_theme_bw() + 
   theme(legend.position="none", 
     panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank(),
    aspect.ratio = 9 / 16,
    plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +  
    labs(x= "Active 2026 ELO", y ="DI Conferences", 
       title = "Current ELOs of NCAA MBB",
       subtitle = "NCAA DI Teams")

g_1
```

```{r}
MAAC_Conference <- merged_mbb_clean %>% 
  filter(conference_2025_26 == 'MAAC')

g_2_MAAC <- ggplot(MAAC_Conference, 
               aes(x = elo_rating, 
                   y = reorder(team_name, elo_rating) 
                   )) + 
  geom_point(alpha = 0.3) + 
 geom_image(aes(image = logo_dark), size = 0.05, asp = 16 / 9) + 
  dark_theme_bw() + 
   theme(legend.position="none", 
     panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank(),
    aspect.ratio = 9 / 16,
    plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +  
    labs(x= "Active 2026 MAAC ELOs", y ="MAAC Teams", 
       title = "Current ELOs of the Metro-Atlantic Athletic Conference",
       )

g_2_MAAC
```

```{r}
ACC_Conference <- merged_mbb_clean %>% 
  filter(conference_2025_26 == 'ACC')

g_3_ACC <- ggplot(ACC_Conference, 
               aes(x = elo_rating, 
                   y = reorder(team_name, elo_rating) 
                   )) + 
  geom_point(alpha = 0.3) + 
 geom_image(aes(image = logo_dark), size = 0.05, asp = 16 / 9) + 
  dark_theme_bw() + 
   theme(legend.position="none", 
     panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank(),
    aspect.ratio = 9 / 16,
    plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +  
    labs(x= "Active 2026 ACC ELOs", y ="ACC Teams", 
       title = "Current ELOs of the Atlantic Coast Conference",
       )

g_3_ACC
```

```{r}
B1G_Conference <- merged_mbb_clean %>% 
  filter(conference_2025_26 == 'Big Ten')

g_4_B1G <- ggplot(B1G_Conference, 
               aes(x = elo_rating, 
                   y = reorder(team_name, elo_rating) 
                   )) + 
  geom_point(alpha = 0.3) + 
 geom_image(aes(image = logo_dark), size = 0.05, asp = 16 / 9) + 
  dark_theme_bw() + 
   theme(legend.position="none", 
     panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank(),
    aspect.ratio = 9 / 16,
    plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +  
    labs(x= "Active 2026 B1G ELOs", y ="Big Ten Teams", 
       title = "Current ELOs of the Big Ten Conference",
       )

g_4_B1G
```

```{r}
A10_Conference <- merged_mbb_clean %>% 
  filter(conference_2025_26 == 'Atlantic 10')

g_5_A10 <- ggplot(A10_Conference, 
               aes(x = elo_rating, 
                   y = reorder(team_name, elo_rating) 
                   )) + 
  geom_point(alpha = 0.3) + 
 geom_image(aes(image = logo_dark), size = 0.05, asp = 16 / 9) + 
  dark_theme_bw() + 
   theme(legend.position="none", 
     panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank(),
    aspect.ratio = 9 / 16,
    plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +  
    labs(x= "Active 2026 A10 ELOs", y ="Atlantic 10 Teams", 
       title = "Current ELOs of the Atlantic 10 Conference",
       )

g_5_A10
```
WORK WITH MAAC CONFERENCE
Testing out probability of win 
Creating a remaining schedule for the MAAC teams based on the Schedule_26 DataFrame

```{r}
# Win Probability of Siena against Saint Peter's example
MAAC_Conference[MAAC_Conference$team_name %in% c("Siena", "Saint Peter's"),]

elo.prob(MAAC_Conference$elo_rating[MAAC_Conference$team_name == "Siena"],
         MAAC_Conference$elo_rating[MAAC_Conference$team_name == "Saint Peter's"])
```

```{r}
# Set seed
set.seed(3)

# Calculate win probability for United Siena v Saint Peter's
win_prob <- elo.prob(MAAC_Conference$elo_rating[MAAC_Conference$team_name == "Siena"],
         MAAC_Conference$elo_rating[MAAC_Conference$team_name == "Saint Peter's"])

# Generate number between 0 and 1
sim <- runif(1, min = 0, max = 1)
print(paste("Siena win probability is:", round(win_prob, 4), sep = " "))
print(paste("The simulated value is:", round(sim, 4), sep = " "))
if(sim < win_prob){
  print(paste("Siena Wins the Game"))
} else{
   print(paste("Saint Peter's Wins the Game"))
}
```

```{r}
MAAC_remaining_schedule <- schedule_26 %>% 
  filter(status_type_name == "STATUS_SCHEDULED") %>% 
  dplyr::rename(
    home_team_name = home_location,
    away_team_name = away_location
  ) %>% 
  filter(home_team_name %in% MAAC_Conference$team_name | 
         away_team_name %in% MAAC_Conference$team_name) %>%
  select(id, game_date, home_team_name, away_team_name)

head(MAAC_remaining_schedule)
```

```{r}
MAAC_remaining_schedule <- merge(
  MAAC_remaining_schedule, 
  MAAC_Conference[, c("team_name", "elo_rating")], 
  by.x = "home_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)

colnames(MAAC_remaining_schedule)[colnames(MAAC_remaining_schedule) == "elo_rating"] <- "home_elo"


MAAC_remaining_schedule
```
Adding in the home team elos in the code above. This will help to build out the remaining schedule and then to build the game predictions/outcomes.

```{r}
MAAC_remaining_schedule <- merge(
  MAAC_remaining_schedule, 
  MAAC_Conference[, c("team_name", "elo_rating")], 
  by.x = "away_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)
colnames(MAAC_remaining_schedule)[colnames(MAAC_remaining_schedule) == "elo_rating"] <- "away_elo"

MAAC_remaining_schedule
```
Getting the win probability and then adding some points to the elo for home court advantage

```{r}
MAAC_remaining_schedule$home_prob <- elo.prob(
  elo.A = MAAC_remaining_schedule$home_elo + 100, 
  elo.B = MAAC_remaining_schedule$away_elo
)
```

Running code similar to what we did for our elo_world_cup.rmd 

```{r}
set.seed(33) 

# Generate a random number (0-1) for every game
MAAC_remaining_schedule$sim_value <- runif(nrow(MAAC_remaining_schedule))

# Determine the winner: If random number < home_prob, Home wins. Else, Away wins.
MAAC_remaining_schedule$predicted_winner <- ifelse(
  MAAC_remaining_schedule$sim_value < MAAC_remaining_schedule$home_prob, 
  MAAC_remaining_schedule$home_team_name, 
  MAAC_remaining_schedule$away_team_name
)

# View the results
head(MAAC_remaining_schedule)
```
Here I am looking at the predicted winners of the remaining MAAC Schedule and how many games these teams are going to win.

```{r}
table(MAAC_remaining_schedule$predicted_winner)
```

```{r}
MAAC_current_elos <- setNames(MAAC_Conference$elo_rating, MAAC_Conference$team_name)

# 2. Sort schedule by date
# It is critical to simulate in order!
MAAC_remaining_schedule <- MAAC_remaining_schedule %>% arrange(game_date)

# 3. Create columns to store results
MAAC_remaining_schedule$home_prob <- NA
MAAC_remaining_schedule$sim_value <- NA
MAAC_remaining_schedule$predicted_winner <- NA
MAAC_remaining_schedule$home_elo_before <- NA
MAAC_remaining_schedule$away_elo_before <- NA

# 4. Loop through every game to simulate and update
# Set K-factor for updates (e.g., 30)
K <- 30
set.seed(33)
```

```{r}
for(i in 1:nrow(MAAC_remaining_schedule)) {
  
  # A. Get Teams
  home_team <- MAAC_remaining_schedule$home_team_name[i]
  away_team <- MAAC_remaining_schedule$away_team_name[i]
  
  # B. Get Current ELOs (Handle missing teams if necessary)
  # If a team isn't in our list (e.g. non-D1), default to 1500 or skip
  h_elo <- if(home_team %in% names(MAAC_current_elos)) MAAC_current_elos[[home_team]] else 1500
  a_elo <- if(away_team %in% names(MAAC_current_elos)) MAAC_current_elos[[away_team]] else 1500
  
  # Store 'Before' ratings for reference
  MAAC_remaining_schedule$home_elo_before[i] <- h_elo
  MAAC_remaining_schedule$away_elo_before[i] <- a_elo
  
  # C. Calculate Win Prob (Current H_ELO + 100 vs Current A_ELO)
  h_prob <- elo.prob(h_elo + 100, a_elo)
  MAAC_remaining_schedule$home_prob[i] <- h_prob
  
  # D. Simulate Outcome
  sim_val <- runif(1)
  MAAC_remaining_schedule$sim_value[i] <- sim_val
  
  if(sim_val < h_prob) {
    # Home Wins
    MAAC_remaining_schedule$predicted_winner[i] <- home_team
    actual_score <- 1
  } else {
    # Away Wins
    MAAC_remaining_schedule$predicted_winner[i] <- away_team
    actual_score <- 0
  }

  # E. UPDATE ELOs!
  # Formula: New = Old + K * (Actual - Expected)
  # Note: The expected result for update excludes the 100 point HCA usually, 
  # or you can include it. Standard packages usually update based on the probability used.
  
  shift <- K * (actual_score - h_prob)
  
  # Apply updates to the lookup table
  if(home_team %in% names(MAAC_current_elos)) {
    MAAC_current_elos[[home_team]] <- h_elo + shift
  }
  if(away_team %in% names(MAAC_current_elos)) {
    MAAC_current_elos[[away_team]] <- a_elo - shift
  }
}

# 5. View the final "Simulated" Standings
final_MAAC_simulated_elos <- data.frame(
  team = names(MAAC_current_elos),
  final_elo = as.numeric(MAAC_current_elos)
) %>% arrange(desc(final_elo))

head(final_MAAC_simulated_elos)
head(MAAC_remaining_schedule)
```

Atlantic 10 Conference

```{r}
# 1. Create a fresh A10 schedule from the FULL schedule_26
A10_remaining_schedule <- schedule_26 %>% 
  filter(status_type_name == "STATUS_SCHEDULED") %>% 
  dplyr::rename(
    home_team_name = home_location,
    away_team_name = away_location
  ) %>% 
  # Filter specifically for A10 teams this time!
  filter(home_team_name %in% A10_Conference$team_name | 
         away_team_name %in% A10_Conference$team_name) %>%
  select(id, game_date, home_team_name, away_team_name)

# 2. Now merge your A10 ELOs onto this fresh schedule
A10_remaining_schedule <- merge(
  A10_remaining_schedule, 
  A10_Conference[, c("team_name", "elo_rating")], 
  by.x = "home_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)

# Rename the column
colnames(A10_remaining_schedule)[colnames(A10_remaining_schedule) == "elo_rating"] <- "home_elo"

# View the results
head(A10_remaining_schedule)

```


```{r}
A10_remaining_schedule <- merge(
  A10_remaining_schedule, 
  A10_Conference[, c("team_name", "elo_rating")], 
  by.x = "away_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)
colnames(A10_remaining_schedule)[colnames(A10_remaining_schedule) == "elo_rating"] <- "away_elo"

A10_remaining_schedule
```

```{r}
A10_remaining_schedule$home_prob <- elo.prob(
  elo.A = A10_remaining_schedule$home_elo + 100, 
  elo.B = A10_remaining_schedule$away_elo
)
```

```{r}
set.seed(33) 

# Generate a random number (0-1) for every game
A10_remaining_schedule$sim_value <- runif(nrow(A10_remaining_schedule))

# Determine the winner: If random number < home_prob, Home wins. Else, Away wins.
A10_remaining_schedule$predicted_winner <- ifelse(
  A10_remaining_schedule$sim_value < A10_remaining_schedule$home_prob, 
  A10_remaining_schedule$home_team_name, 
  A10_remaining_schedule$away_team_name
)

# View the results
head(A10_remaining_schedule)
```

```{r}
A10_current_elos <- setNames(A10_Conference$elo_rating, A10_Conference$team_name)

# 2. Sort schedule by date
# It is critical to simulate in order!
A10_remaining_schedule <- A10_remaining_schedule %>% arrange(game_date)

# 3. Create columns to store results
A10_remaining_schedule$home_prob <- NA
A10_remaining_schedule$sim_value <- NA
A10_remaining_schedule$predicted_winner <- NA
A10_remaining_schedule$home_elo_before <- NA
A10_remaining_schedule$away_elo_before <- NA

# 4. Loop through every game to simulate and update
# Set K-factor for updates (e.g., 30)
K <- 30
set.seed(33)
```

```{r}
for(i in 1:nrow(A10_remaining_schedule)) {
  
  # A. Get Teams
  home_team <- A10_remaining_schedule$home_team_name[i]
  away_team <- A10_remaining_schedule$away_team_name[i]
  
  # B. Get Current ELOs (Handle missing teams if necessary)
  # If a team isn't in our list (e.g. non-D1), default to 1500 or skip
  h_elo <- if(home_team %in% names(A10_current_elos)) A10_current_elos[[home_team]] else 1500
  a_elo <- if(away_team %in% names(A10_current_elos)) A10_current_elos[[away_team]] else 1500
  
  # Store 'Before' ratings for reference
  A10_remaining_schedule$home_elo_before[i] <- h_elo
  A10_remaining_schedule$away_elo_before[i] <- a_elo
  
  # C. Calculate Win Prob (Current H_ELO + 100 vs Current A_ELO)
  h_prob <- elo.prob(h_elo + 100, a_elo)
  A10_remaining_schedule$home_prob[i] <- h_prob
  
  # D. Simulate Outcome
  sim_val <- runif(1)
  A10_remaining_schedule$sim_value[i] <- sim_val
  
  if(sim_val < h_prob) {
    # Home Wins
    A10_remaining_schedule$predicted_winner[i] <- home_team
    actual_score <- 1
  } else {
    # Away Wins
    A10_remaining_schedule$predicted_winner[i] <- away_team
    actual_score <- 0
  }

  # E. UPDATE ELOs!
  # Formula: New = Old + K * (Actual - Expected)
  # Note: The expected result for update excludes the 100 point HCA usually, 
  # or you can include it. Standard packages usually update based on the probability used.
  
  shift <- K * (actual_score - h_prob)
  
  # Apply updates to the lookup table
  if(home_team %in% names(A10_current_elos)) {
    A10_current_elos[[home_team]] <- h_elo + shift
  }
  if(away_team %in% names(A10_current_elos)) {
    A10_current_elos[[away_team]] <- a_elo - shift
  }
}

# 5. View the final "Simulated" Standings
A10_final_simulated_elos <- data.frame(
  team = names(A10_current_elos),
  A10_final_elo = as.numeric(A10_current_elos)
) %>% arrange(desc(A10_final_elo))

head(A10_final_simulated_elos)
head(A10_remaining_schedule)
```
Big Ten Conference Sims 

```{r}
B1G_remaining_schedule <- schedule_26 %>% 
  filter(status_type_name == "STATUS_SCHEDULED") %>% 
  dplyr::rename(
    home_team_name = home_location,
    away_team_name = away_location
  ) %>% 
  # Filter specifically for A10 teams this time!
  filter(home_team_name %in% B1G_Conference$team_name | 
         away_team_name %in% B1G_Conference$team_name) %>%
  select(id, game_date, home_team_name, away_team_name)

# 2. Now merge your A10 ELOs onto this fresh schedule
B1G_remaining_schedule <- merge(
  B1G_remaining_schedule, 
  B1G_Conference[, c("team_name", "elo_rating")], 
  by.x = "home_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)

# Rename the column
colnames(B1G_remaining_schedule)[colnames(B1G_remaining_schedule) == "elo_rating"] <- "home_elo"

# View the results
head(B1G_remaining_schedule)

```
```{r}
B1G_remaining_schedule <- merge(
  B1G_remaining_schedule, 
  B1G_Conference[, c("team_name", "elo_rating")], 
  by.x = "away_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)
colnames(B1G_remaining_schedule)[colnames(B1G_remaining_schedule) == "elo_rating"] <- "away_elo"

B1G_remaining_schedule
```
```{r}
B1G_remaining_schedule$home_prob <- elo.prob(
  elo.A = B1G_remaining_schedule$home_elo + 100, 
  elo.B = B1G_remaining_schedule$away_elo
)
```

```{r}
set.seed(33) 

# Generate a random number (0-1) for every game
B1G_remaining_schedule$sim_value <- runif(nrow(B1G_remaining_schedule))

# Determine the winner: If random number < home_prob, Home wins. Else, Away wins.
B1G_remaining_schedule$predicted_winner <- ifelse(
  B1G_remaining_schedule$sim_value < B1G_remaining_schedule$home_prob, 
  B1G_remaining_schedule$home_team_name, 
  B1G_remaining_schedule$away_team_name
)

# View the results
head(B1G_remaining_schedule)
```

```{r}
B1G_current_elos <- setNames(B1G_Conference$elo_rating, B1G_Conference$team_name)

# 2. Sort schedule by date
# It is critical to simulate in order!
B1G_remaining_schedule <- B1G_remaining_schedule %>% arrange(game_date)

# 3. Create columns to store results
B1G_remaining_schedule$home_prob <- NA
B1G_remaining_schedule$sim_value <- NA
B1G_remaining_schedule$predicted_winner <- NA
B1G_remaining_schedule$home_elo_before <- NA
B1G_remaining_schedule$away_elo_before <- NA

# 4. Loop through every game to simulate and update
# Set K-factor for updates (e.g., 30)
K <- 30
set.seed(33)
```

```{r}
for(i in 1:nrow(B1G_remaining_schedule)) {
  
  # A. Get Teams
  home_team <- B1G_remaining_schedule$home_team_name[i]
  away_team <- B1G_remaining_schedule$away_team_name[i]
  
  # B. Get Current ELOs (Handle missing teams if necessary)
  # If a team isn't in our list (e.g. non-D1), default to 1500 or skip
  h_elo <- if(home_team %in% names(B1G_current_elos)) B1G_current_elos[[home_team]] else 1500
  a_elo <- if(away_team %in% names(B1G_current_elos)) B1G_current_elos[[away_team]] else 1500

  # Store 'Before' ratings for reference
  B1G_remaining_schedule$home_elo_before[i] <- h_elo
  B1G_remaining_schedule$away_elo_before[i] <- a_elo
  
  # C. Calculate Win Prob (Current H_ELO + 100 vs Current A_ELO)
  h_prob <- elo.prob(h_elo + 100, a_elo)
  B1G_remaining_schedule$home_prob[i] <- h_prob
  
  # D. Simulate Outcome
  sim_val <- runif(1)
  B1G_remaining_schedule$sim_value[i] <- sim_val
  
  if(sim_val < h_prob) {
    # Home Wins
    B1G_remaining_schedule$predicted_winner[i] <- home_team
    actual_score <- 1
  } else {
    # Away Wins
    B1G_remaining_schedule$predicted_winner[i] <- away_team
    actual_score <- 0
  }

  # E. UPDATE ELOs!
  # Formula: New = Old + K * (Actual - Expected)
  # Note: The expected result for update excludes the 100 point HCA usually, 
  # or you can include it. Standard packages usually update based on the probability used.
  
  shift <- K * (actual_score - h_prob)
  
  # Apply updates to the lookup table
  if(home_team %in% names(A10_current_elos)) {
    B1G_current_elos[[home_team]] <- h_elo + shift
  }
  if(away_team %in% names(B1G_current_elos)) {
    B1G_current_elos[[away_team]] <- a_elo - shift
  }
}

# 5. View the final "Simulated" Standings
B1G_final_simulated_elos <- data.frame(
  team = names(B1G_current_elos),
  B1G_final_elo = as.numeric(B1G_current_elos)
) %>% arrange(desc(B1G_final_elo))

head(B1G_final_simulated_elos)
head(B1G_remaining_schedule)
```

Atlantic Coast Conference

```{r}
ACC_remaining_schedule <- schedule_26 %>% 
  filter(status_type_name == "STATUS_SCHEDULED") %>% 
  dplyr::rename(
    home_team_name = home_location,
    away_team_name = away_location
  ) %>% 
  # Filter specifically for ACC teams this time
  filter(home_team_name %in% ACC_Conference$team_name | 
         away_team_name %in% ACC_Conference$team_name) %>%
  select(id, game_date, home_team_name, away_team_name)

# 2. Now merge your A10 ELOs onto this fresh schedule
ACC_remaining_schedule <- merge(
  ACC_remaining_schedule, 
  ACC_Conference[, c("team_name", "elo_rating")], 
  by.x = "home_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)

# Rename the column
colnames(ACC_remaining_schedule)[colnames(ACC_remaining_schedule) == "elo_rating"] <- "home_elo"

# View the results
head(ACC_remaining_schedule)
```
```{r}
ACC_remaining_schedule <- merge(
  ACC_remaining_schedule, 
  ACC_Conference[, c("team_name", "elo_rating")], 
  by.x = "away_team_name", 
  by.y = "team_name", 
  all.x = TRUE
)
colnames(ACC_remaining_schedule)[colnames(ACC_remaining_schedule) == "elo_rating"] <- "away_elo"

ACC_remaining_schedule
```
```{r}
ACC_remaining_schedule$home_prob <- elo.prob(
  elo.A = ACC_remaining_schedule$home_elo + 100, 
  elo.B = ACC_remaining_schedule$away_elo
)
```

```{r}
set.seed(33) 

# Generate a random number (0-1) for every game
ACC_remaining_schedule$sim_value <- runif(nrow(ACC_remaining_schedule))

# Determine the winner: If random number < home_prob, Home wins. Else, Away wins
ACC_remaining_schedule$predicted_winner <- ifelse(
  ACC_remaining_schedule$sim_value < ACC_remaining_schedule$home_prob, 
  ACC_remaining_schedule$home_team_name, 
  ACC_remaining_schedule$away_team_name
)

# View the results
head(ACC_remaining_schedule)
```
```{r}
ACC_current_elos <- setNames(ACC_Conference$elo_rating, ACC_Conference$team_name)

# 2. Sort schedule by date
# It is critical to simulate in order
ACC_remaining_schedule <- ACC_remaining_schedule %>% arrange(game_date)

# 3. Create columns to store results
ACC_remaining_schedule$home_prob <- NA
ACC_remaining_schedule$sim_value <- NA
ACC_remaining_schedule$predicted_winner <- NA
ACC_remaining_schedule$home_elo_before <- NA
ACC_remaining_schedule$away_elo_before <- NA

# 4. Loop through every game to simulate and update
# Set K-factor for updates (e.g., 30)
K <- 30
set.seed(33)
```

```{r}
for(i in 1:nrow(ACC_remaining_schedule)) {
  
  # A. Get Teams
  home_team <- ACC_remaining_schedule$home_team_name[i]
  away_team <- ACC_remaining_schedule$away_team_name[i]
  
  # B. Get Current ELOs (Handle missing teams if necessary)
  # If a team isn't in our list (e.g. non-D1), default to 1500 or skip
  h_elo <- if(home_team %in% names(ACC_current_elos)) ACC_current_elos[[home_team]] else 1500
  a_elo <- if(away_team %in% names(ACC_current_elos)) ACC_current_elos[[away_team]] else 1500

  # Store 'Before' ratings for reference
  ACC_remaining_schedule$home_elo_before[i] <- h_elo
  ACC_remaining_schedule$away_elo_before[i] <- a_elo
  
  # C. Calculate Win Prob (Current H_ELO + 100 vs Current A_ELO)
  h_prob <- elo.prob(h_elo + 100, a_elo)
  ACC_remaining_schedule$home_prob[i] <- h_prob
  
  # D. Simulate Outcome
  sim_val <- runif(1)
  ACC_remaining_schedule$sim_value[i] <- sim_val
  
  if(sim_val < h_prob) {
    # Home Wins
    ACC_remaining_schedule$predicted_winner[i] <- home_team
    actual_score <- 1
  } else {
    # Away Wins
    ACC_remaining_schedule$predicted_winner[i] <- away_team
    actual_score <- 0
  }

  # E. UPDATE ELOs!
  # Formula: New = Old + K * (Actual - Expected)
  # Note: The expected result for update excludes the 100 point HCA usually, 
  # or you can include it. Standard packages usually update based on the probability used.
  
  shift <- K * (actual_score - h_prob)
  
  # Apply updates to the lookup table
  if(home_team %in% names(A10_current_elos)) {
    ACC_current_elos[[home_team]] <- h_elo + shift
  }
  if(away_team %in% names(B1G_current_elos)) {
    ACC_current_elos[[away_team]] <- a_elo - shift
  }
}

# 5. View the final "Simulated" Standings
ACC_final_simulated_elos <- data.frame(
  team = names(ACC_current_elos),
  ACC_final_elo = as.numeric(ACC_current_elos)
) %>% arrange(desc(ACC_final_elo))

head(ACC_final_simulated_elos)
head(ACC_remaining_schedule)
```
```{r}
save(espn_mbb_pbp, temp_1, temp_2, games_1, games_2, mbb_2026, mbb_2025, df_elo_25, df_elo_26, conferences_26_clean, schedule_26, ACC_Conference, ACC_final_simulated_elos, ACC_remaining_schedule, A10_Conference, A10_final_simulated_elos, A10_remaining_schedule, B1G_Conference, B1G_final_simulated_elos, B1G_remaining_schedule, MAAC_Conference, MAAC_remaining_schedule, merged_mbb_clean,  file = "mbb_data_3.rda")
```

BUILD OUT THE TOURNAMENT FIELD
WORK WITH MARTIN FOR MORE ACCURATE ELOs


```{r}
 ranking_data <- data.frame(
  Rank = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25),
  Team = c(
    "Purdue", "Houston", "Florida", "UConn", "St. John's",
    "Duke", "Michigan", "BYU", "Kentucky", "Texas Tech",
    "Louisville", "UCLA", "Arizona", "Arkansas", "Alabama",
    "Iowa State", "Illinois", "Tennessee", "Kansas", "Auburn",
    "Gonzaga", "Michigan State", "Creighton", "Wisconsin", "North Carolina"
  ),
  Points = c(
    1485, 1459, 1382, 1299, 1203, 1123, 1084, 1078, 1056, 1015,
    966, 741, 715, 695, 620, 616, 567, 462, 453, 424,
    387, 188, 158, 136, 104
  ) / 2
)

# Print the dataframe to verify
print(ranking_data)
```


```{r}

season_games <- load_mbb_schedule(2026) %>%
  filter(status_type_completed == TRUE) 


elo_results <- elo.run(score(home_score, away_score) ~ home_location + away_location, 
                       data = season_games, 
                       k = 30)

final_elos <- as.data.frame(final.elos(elo_results))

team_vec <- final_elos
team_vec[,1] <- 1500

for(i  in 1:nrow(ranking_data)){
  team_vec[which(rownames(team_vec) == ranking_data$Team[i]),1] <- 1500 + ranking_data$Points[i]
}

team_vec[which(rownames(team_vec) %in% ranking_data$Team),]

team_vals <- team_vec[,1]
names(team_vals) <- rownames(team_vec)

elo_results <- elo.run(score(home_score, away_score) ~ home_location + away_location, 
                       data = season_games[2,], 
                       k = 30, initial.elos = team_vals)



final_elos <- as.data.frame(final.elos(elo_results))
colnames(final_elos) <- "projected_elo"

team_elo_res <- team_vals

k_vec <- rep(20, nrow(season_games))

k_vec[which(season_games$home_conference_id == season_games$away_conference_id)] <- 25


k_vec[which(season_games$home_conference_id == "3")] <- 20
k_vec[which(season_games$home_conference_id == "2")] <- 30
k_vec[which(season_games$home_conference_id == "1")] <- 15
k_vec[which(season_games$home_conference_id == "62")] <- 18
k_vec[which(season_games$home_conference_id == "46")] <- 15
k_vec[which(season_games$home_conference_id == "8")] <- 30
k_vec[which(season_games$home_conference_id == "4")] <- 25
k_vec[which(season_games$home_conference_id == "5")] <- 15
k_vec[which(season_games$home_conference_id == "6")] <- 15
k_vec[which(season_games$home_conference_id == "7")] <- 30
k_vec[which(season_games$home_conference_id == "9")] <- 18
k_vec[which(season_games$home_conference_id == "10")] <- 15
k_vec[which(season_games$home_conference_id == "11")] <- 15
k_vec[which(season_games$home_conference_id == "45")] <- 15
k_vec[which(season_games$home_conference_id == "12")] <- 15
k_vec[which(season_games$home_conference_id == "13")] <- 20
k_vec[which(season_games$home_conference_id == "14")] <- 18
k_vec[which(season_games$home_conference_id == "16")] <- 15
k_vec[which(season_games$home_conference_id == "44")] <- 20
k_vec[which(season_games$home_conference_id == "18")] <- 15
k_vec[which(season_games$home_conference_id == "19")] <- 15
k_vec[which(season_games$home_conference_id == "20")] <- 15
k_vec[which(season_games$home_conference_id == "23")] <- 30
k_vec[which(season_games$home_conference_id == "24")] <- 15
k_vec[which(season_games$home_conference_id == "25")] <- 18
k_vec[which(season_games$home_conference_id == "49")] <- 15
k_vec[which(season_games$home_conference_id == "27")] <- 18
k_vec[which(season_games$home_conference_id == "26")] <- 15
k_vec[which(season_games$home_conference_id == "30")] <- 15
k_vec[which(season_games$home_conference_id == "29")] <- 20

season_games <- as.data.frame(season_games)
for(i in 1:nrow(season_games)) {
  temp <- elo.run(score(home_score, away_score) ~ home_location + away_location, 
                       data = season_games[i,], 
                       k = k_vec[i], 
                       initial.elos = team_elo_res)
  t1 <- as.data.frame(temp)
  team_elo_res[which(names(team_elo_res) == t1$team.A)] <- t1$elo.A
  team_elo_res[which(names(team_elo_res) == t1$team.B)] <- t1$elo.B
  #final_elos <- as.data.frame(final.elos(temp))
  #team_elo_res <- final_elos[,1]
#names(team_elo_res) <- rownames(final_elos)
  
}

final_elos <- as.data.frame(team_elo_res)
projected_standings <- final_elos
projected_standings$team <- rownames(final_elos)  
rownames(projected_standings) <- NULL             


projected_standings <- projected_standings[order(projected_standings$team_elo_res, decreasing = TRUE), ]


head(projected_standings)
```
Build out the different conference tournaments. Most tournaments follow one of three types of sturcutes which is reflected in the code chunk below with examples for each.

```{r}
# --- PREP CONFERENCE DATA ---
final_standings_conf <- merge(
  x = projected_standings,
  y = conferences_26_clean,
  by.x = "team",
  by.y = "team",
  all.x = TRUE
) %>% 
  dplyr::rename(elo_rating = team_elo_res) %>%
  drop_na(conference_2025_26)

# Helper function to simulate a single game
sim_game <- function(t_a, t_b, conf_data) {
  prob <- elo.prob(conf_data$elo_rating[conf_data$team == t_a], 
                   conf_data$elo_rating[conf_data$team == t_b])
  ifelse(runif(1) < prob, t_a, t_b)
}


# --- THE 4 Tournament Structures for 2026 ---

# Traditional Single-Elimination (No Byes)
# Handles standard 8-team brackets, and the 4-team Ivy League bracket
sim_traditional_tourney <- function(conf_data, is_ivy = FALSE) {
  
  if(is_ivy) {
    teams <- conf_data %>% arrange(desc(elo_rating)) %>% head(4)
    if(nrow(teams) < 4) return(list(Champion = teams$team[1]))
    
    # Ivy League is just Semis and Finals
    sf1 <- sim_game(teams$team[1], teams$team[4], conf_data)
    sf2 <- sim_game(teams$team[2], teams$team[3], conf_data)
    champ <- sim_game(sf1, sf2, conf_data)
    
    return(list(Semifinals = c(sf1, sf2), Champion = champ))
  } 
  
  else {
    teams <- conf_data %>% arrange(desc(elo_rating)) %>% head(8)
    if(nrow(teams) < 8) return(list(Champion = teams$team[1]))
    
    # Quarterfinals
    qf1 <- sim_game(teams$team[1], teams$team[8], conf_data)
    qf2 <- sim_game(teams$team[4], teams$team[5], conf_data)
    qf3 <- sim_game(teams$team[3], teams$team[6], conf_data)
    qf4 <- sim_game(teams$team[2], teams$team[7], conf_data)
    
    # Semifinals
    sf1 <- sim_game(qf1, qf2, conf_data)
    sf2 <- sim_game(qf3, qf4, conf_data)
    
    # Championship
    champ <- sim_game(sf1, sf2, conf_data)
    
    return(list(Quarterfinals = c(qf1, qf2, qf3, qf4), Semifinals = c(sf1, sf2), Champion = champ))
  }
}

# 2. Single-Bye Format (Standard Staggered)
# Top seeds get a bye to Quarter Finals, lower seeds play in R1
sim_single_bye_tourney <- function(conf_data) {
  teams <- conf_data %>% arrange(desc(elo_rating)) %>% head(10)
  if(nrow(teams) < 10) return(list(Champion = teams$team[1])) 
  
  # Round 1 (Bottom seeds play-in)
  r1_g1 <- sim_game(teams$team[8], teams$team[9], conf_data)
  r1_g2 <- sim_game(teams$team[7], teams$team[10], conf_data)
  
  # Quarterfinals (Top seeds + R1 Winners)
  qf_1 <- sim_game(teams$team[1], r1_g1, conf_data)         
  qf_2 <- sim_game(teams$team[4], teams$team[5], conf_data) 
  qf_3 <- sim_game(teams$team[2], r1_g2, conf_data)         
  qf_4 <- sim_game(teams$team[3], teams$team[6], conf_data) 
  
  sf_1 <- sim_game(qf_1, qf_2, conf_data)
  sf_2 <- sim_game(qf_3, qf_4, conf_data)
  
  champ <- sim_game(sf_1, sf_2, conf_data)
  
  return(list(Round_1 = c(r1_g1, r1_g2), Quarterfinals = c(qf_1, qf_2, qf_3, qf_4), Semifinals = c(sf_1, sf_2), Champion = champ))
}

# 3. Double-Bye & Triple-Bye Format (The Power Model)
# Top 4 seeds jump straight to QF (3 wins to a title). 
sim_power_tourney <- function(conf_data) {
  teams <- conf_data %>% arrange(desc(elo_rating)) %>% head(15)
  if(nrow(teams) < 15) return(list(Champion = teams$team[1]))
  
  # Round 1
  r1_g1 <- sim_game(teams$team[12], teams$team[13], conf_data)
  r1_g2 <- sim_game(teams$team[11], teams$team[14], conf_data)
  r1_g3 <- sim_game(teams$team[10], teams$team[15], conf_data)
  
  # Round 2
  r2_g1 <- sim_game(teams$team[8], teams$team[9], conf_data)
  r2_g2 <- sim_game(teams$team[5], r1_g1, conf_data)
  r2_g3 <- sim_game(teams$team[6], r1_g2, conf_data)
  r2_g4 <- sim_game(teams$team[7], r1_g3, conf_data)
  
  # Quarterfinals (Top 4 enter here)
  qf_1 <- sim_game(teams$team[1], r2_g1, conf_data)
  qf_2 <- sim_game(teams$team[4], r2_g2, conf_data)
  qf_3 <- sim_game(teams$team[2], r2_g4, conf_data)
  qf_4 <- sim_game(teams$team[3], r2_g3, conf_data)
  
  sf_1 <- sim_game(qf_1, qf_2, conf_data)
  sf_2 <- sim_game(qf_3, qf_4, conf_data)
  
  champ <- sim_game(sf_1, sf_2, conf_data)
  
  return(list(Round_1 = c(r1_g1, r1_g2, r1_g3), Round_2 = c(r2_g1, r2_g2, r2_g3, r2_g4), Quarterfinals = c(qf_1, qf_2, qf_3, qf_4), Semifinals = c(sf_1, sf_2), Champion = champ))
}

# 4. The Stepladder Format (Direct to Semifinals)
# Top 2 seeds jump straight to SF (2 wins to a title).
sim_stepladder_tourney <- function(conf_data) {
  teams <- conf_data %>% arrange(desc(elo_rating)) %>% head(8)
  if(nrow(teams) < 8) return(list(Champion = teams$team[1])) 
  
  # Round 1
  r1_1 <- sim_game(teams$team[5], teams$team[8], conf_data)
  r1_2 <- sim_game(teams$team[6], teams$team[7], conf_data)
  
  # Quarterfinals
  qf_1 <- sim_game(teams$team[4], r1_1, conf_data)
  qf_2 <- sim_game(teams$team[3], r1_2, conf_data)
  
  # Semifinals (Top 2 enter here)
  sf_1 <- sim_game(teams$team[1], qf_1, conf_data)
  sf_2 <- sim_game(teams$team[2], qf_2, conf_data)
  
  champ <- sim_game(sf_1, sf_2, conf_data)
  
  return(list(Round_1 = c(r1_1, r1_2), Quarterfinals = c(qf_1, qf_2), Semifinals = c(sf_1, sf_2), Champion = champ))
}


# --- RUN TOURNAMENT SIMULATIONS ---

conf_split <- split(final_standings_conf, final_standings_conf$conference_2025_26)
auto_bids <- character(0)
all_tourney_results <- list() 

# Define the groupings 
stepladder_confs <- c("WCC", "OVC", "Southland")
power_confs <- c("AAC", "Atlantic 10", "ACC", "Big 12", "CAA", "SEC", "Sun Belt", "Big Ten")
single_bye_confs <- c("ASUN", "Big East", "Big Sky", "Big South", "Horizon", "MAAC", "MVC", "Mountain West", "Patriot", "SoCon")
# The remaining conferences default to Traditional Single-Elimination

for(conf_name in names(conf_split)) {
  conf_data <- conf_split[[conf_name]]
  
  # Route to the correct mathematical bracket
  if(conf_name %in% stepladder_confs) {
    results <- sim_stepladder_tourney(conf_data)
  } 
  else if(conf_name %in% power_confs) {
    results <- sim_power_tourney(conf_data)
  } 
  else if(conf_name %in% single_bye_confs) {
    results <- sim_single_bye_tourney(conf_data)
  } 
  else if(conf_name == "Ivy League") {
    # Ivy League uses Traditional but only invites 4 teams
    results <- sim_traditional_tourney(conf_data, is_ivy = TRUE)
  } 
  else {
    # Default: Traditional 8-team bracket
    results <- sim_traditional_tourney(conf_data, is_ivy = FALSE)
  }
  
  all_tourney_results[[conf_name]] <- results
  auto_bids <- c(auto_bids, results$Champion)
}

# --- 4. ASSEMBLE THE BIDS ---

auto_bid_df <- final_standings_conf %>% filter(team %in% auto_bids)
at_large_df <- final_standings_conf %>% filter(!team %in% auto_bids) %>% arrange(desc(elo_rating)) %>% head(36)

# --- 5. THE FIRST FOUR (PLAY-INS) ---

auto_locks <- head(auto_bid_df %>% arrange(desc(elo_rating)), nrow(auto_bid_df) - 4)
auto_playin <- tail(auto_bid_df %>% arrange(desc(elo_rating)), 4)

at_large_locks <- head(at_large_df, nrow(at_large_df) - 4)
at_large_playin <- tail(at_large_df, 4)

play_playin <- function(t1, t2) {
  prob <- elo.prob(t1$elo_rating, t2$elo_rating)
  if(runif(1) < prob) return(t1) else return(t2)
}

ff_w1 <- play_playin(auto_playin[1,], auto_playin[2,])
ff_w2 <- play_playin(auto_playin[3,], auto_playin[4,])
ff_w3 <- play_playin(at_large_playin[1,], at_large_playin[2,])
ff_w4 <- play_playin(at_large_playin[3,], at_large_playin[4,])



final_64_unranked <- bind_rows(auto_locks, at_large_locks, ff_w1, ff_w2, ff_w3, ff_w4)

```


```{r}

top_64 <- final_64_unranked %>%
  arrange(desc(elo_rating)) %>%
  head(64) %>%
  dplyr::rename(team_name = team) %>%
  mutate(overall_rank = 1:64)


regions <- rep(c("East", "West", "South", "Midwest"), 16)

# Snake the draft for competitive balance
for(i in 1:16) {
  if(i %% 2 == 0) {
    indices <- ((i-1)*4 + 1):(i*4)
    regions[indices] <- rev(regions[indices])
  }
}

top_64$region <- regions
top_64$regional_seed <- rep(1:16, each = 4)


bracket_order <- c(1, 16, 8, 9, 5, 12, 4, 13, 6, 11, 3, 14, 7, 10, 2, 15)

tourney_field <- data.frame() 

# Loop through regions to sort teams into correct Bracket Order
for(r in unique(regions)) {
  # Get teams for this region
  region_teams <- top_64[top_64$region == r, ]
  
  # Sort them: 1 vs 16, 8 vs 9, etc.
  region_teams <- region_teams[match(bracket_order, region_teams$regional_seed), ]
  
  # Add to the final dataframe
  tourney_field <- rbind(tourney_field, region_teams)
}

print(head(tourney_field))

```


```{r}
# 1. Grab all "Top" teams in the bracket (Rows 1, 3, 5, etc.)
team_A <- tourney_field[seq(1, 64, by = 2), ]

# 2. Grab all "Bottom" teams in the bracket (Rows 2, 4, 6, etc.)
team_B <- tourney_field[seq(2, 64, by = 2), ]

# 3. Combine them into a single Matchup Dataframe
r64_matchups <- data.frame(
  Region = team_A$region,
  Seed_A = team_A$regional_seed,
  Team_A = team_A$team_name,
  Elo_A  = team_A$elo_rating,
  
  # "VS" column just for readability
  VS = "vs",
  
  Seed_B = team_B$regional_seed,
  Team_B = team_B$team_name,
  Elo_B  = team_B$elo_rating
)

# 4. View the matchups to confirm logic (Check that 1 is playing 16)
head(r64_matchups, 10)
```

```{r}

# --- 1. SETUP & INITIALIZATION ---

set.seed(123452033)
k <- 30

# Initialize columns to track progress through the tournament
# We assume 'tourney_field' is already sorted by matchup order (1v16, 8v9, etc.)
tourney_field$r64_game <- rep(1:32, each = 2) # Assign Game IDs 1-32
tourney_field$r32_game <- NA
tourney_field$r16_game <- NA
tourney_field$r8_game  <- NA
tourney_field$r4_game  <- NA
tourney_field$r2_game  <- NA
tourney_field$champion <- NA

# --- 2. ROUND OF 64 (32 Games) ---
cat("\n\n=== ROUND OF 64 ===\n")

for(i in 1:32){
  # Find the 2 teams playing in Game 'i'
  indices <- which(tourney_field$r64_game == i)
  match_teams <- tourney_field[indices, ]
  
  # Calculate Win Probability (No Home Court)
  prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
  sim <- runif(1)
  
  if(sim < prob) {
    # Team 1 Wins
    winner_idx <- indices[1]
    res <- 1
    cat(sprintf("Game %02d: %-20s def. %-20s\n", i, match_teams$team_name[1], match_teams$team_name[2]))
  } else {
    # Team 2 Wins (Upset?)
    winner_idx <- indices[2]
    res <- 0
    cat(sprintf("Game %02d: %-20s def. %-20s\n", i, match_teams$team_name[2], match_teams$team_name[1]))
  }
  
  # Update Elos
  new_elo <- elo.calc(wins.A = res, elo.A = match_teams$elo_rating[1], elo.B = match_teams$elo_rating[2], k = k)
  tourney_field$elo_rating[indices] <- as.numeric(new_elo[1, ])
  
  # Advance Winner to Next Round (Game ID maps to ceiling(i/2))
  tourney_field$r32_game[winner_idx] <- ceiling(i / 2)
}

# --- 3. ROUND OF 32 (16 Games) ---
cat("\n\n=== ROUND OF 32 ===\n")

for(i in 1:16){
  indices <- which(tourney_field$r32_game == i)
  match_teams <- tourney_field[indices, ]
  
  # Safety check: ensure 2 teams made it
  if(length(indices) == 2) {
    prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
    sim <- runif(1)
    
    if(sim < prob) {
      winner_idx <- indices[1]
      res <- 1
      cat(sprintf("Game %02d: %-20s def. %-20s\n", i, match_teams$team_name[1], match_teams$team_name[2]))
    } else {
      winner_idx <- indices[2]
      res <- 0
      cat(sprintf("Game %02d: %-20s def. %-20s\n", i, match_teams$team_name[2], match_teams$team_name[1]))
    }
    
    new_elo <- elo.calc(wins.A = res, elo.A = match_teams$elo_rating[1], elo.B = match_teams$elo_rating[2], k = k)
    tourney_field$elo_rating[indices] <- as.numeric(new_elo[1, ])
    
    tourney_field$r16_game[winner_idx] <- ceiling(i / 2)
  }
}

# --- 4. SWEET 16 (8 Games) ---
cat("\n\n=== SWEET 16 ===\n")

for(i in 1:8){
  indices <- which(tourney_field$r16_game == i)
  match_teams <- tourney_field[indices, ]
  
  if(length(indices) == 2) {
    prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
    sim <- runif(1)
    
    if(sim < prob) {
      winner_idx <- indices[1]
      res <- 1
      cat(sprintf("Game %d: %-20s def. %-20s\n", i, match_teams$team_name[1], match_teams$team_name[2]))
    } else {
      winner_idx <- indices[2]
      res <- 0
      cat(sprintf("Game %d: %-20s def. %-20s\n", i, match_teams$team_name[2], match_teams$team_name[1]))
    }
    
    new_elo <- elo.calc(wins.A = res, elo.A = match_teams$elo_rating[1], elo.B = match_teams$elo_rating[2], k = k)
    tourney_field$elo_rating[indices] <- as.numeric(new_elo[1, ])
    
    tourney_field$r8_game[winner_idx] <- ceiling(i / 2)
  }
}

# --- 5. ELITE 8 (4 Games) ---
cat("\n\n=== ELITE 8 ===\n")

for(i in 1:4){
  indices <- which(tourney_field$r8_game == i)
  match_teams <- tourney_field[indices, ]
  
  if(length(indices) == 2) {
    prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
    sim <- runif(1)
    
    if(sim < prob) {
      winner_idx <- indices[1]
      res <- 1
      cat(sprintf("Region %d: %-20s def. %-20s\n", i, match_teams$team_name[1], match_teams$team_name[2]))
    } else {
      winner_idx <- indices[2]
      res <- 0
      cat(sprintf("Region %d: %-20s def. %-20s\n", i, match_teams$team_name[2], match_teams$team_name[1]))
    }
    
    new_elo <- elo.calc(wins.A = res, elo.A = match_teams$elo_rating[1], elo.B = match_teams$elo_rating[2], k = k)
    tourney_field$elo_rating[indices] <- as.numeric(new_elo[1, ])
    
    tourney_field$r4_game[winner_idx] <- ceiling(i / 2)
  }
}

# --- 6. FINAL FOUR (2 Games) ---
cat("\n\n=== FINAL FOUR ===\n")

for(i in 1:2){
  indices <- which(tourney_field$r4_game == i)
  match_teams <- tourney_field[indices, ]
  
  if(length(indices) == 2) {
    prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
    sim <- runif(1)
    
    if(sim < prob) {
      winner_idx <- indices[1]
      res <- 1
      cat(sprintf("Semi %d: %-20s def. %-20s\n", i, match_teams$team_name[1], match_teams$team_name[2]))
    } else {
      winner_idx <- indices[2]
      res <- 0
      cat(sprintf("Semi %d: %-20s def. %-20s\n", i, match_teams$team_name[2], match_teams$team_name[1]))
    }
    
    new_elo <- elo.calc(wins.A = res, elo.A = match_teams$elo_rating[1], elo.B = match_teams$elo_rating[2], k = k)
    tourney_field$elo_rating[indices] <- as.numeric(new_elo[1, ])
    
    tourney_field$r2_game[winner_idx] <- 1 # Both winners go to Game 1 (The Championship)
  }
}

# --- 7. NATIONAL CHAMPIONSHIP (1 Game) ---
cat("\n\n=== NATIONAL CHAMPIONSHIP ===\n")

indices <- which(tourney_field$r2_game == 1)
match_teams <- tourney_field[indices, ]

if(length(indices) == 2) {
  prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
  sim <- runif(1)
  
  if(sim < prob) {
    cat(sprintf("CHAMPION: %s\n", match_teams$team_name[1]))
    tourney_field$champion[indices[1]] <- 1
  } else {
    cat(sprintf("CHAMPION: %s\n", match_teams$team_name[2]))
    tourney_field$champion[indices[2]] <- 1
  }
}
```

```{r}
# Set number of simulations
n_sims <- 10000

# Set seed for reproducibility
set.seed(1234520733)

# Create empty list to store results
results_list <- list()

# Define K-factor for dynamic Elo updates inside the tournament
k <- 30

# --- START SIMULATION LOOP ---
for(s in 1:n_sims){
  
  # 1. Reset team database to initial state using your qualified teams
  # We assume 'tourney_field' is already sorted by bracket matchup (1vs16, 8vs9, etc.)
  team_db_sim <- tourney_field
  
  # 2. Initialize columns for Game IDs
  # Round of 64: Assign games 1-32 based on row order
  team_db_sim$games_64 <- rep(1:32, each = 2) 
  
  # Future rounds: Initialize as NA
  team_db_sim$games_32 <- NA
  team_db_sim$games_16 <- NA
  team_db_sim$games_8  <- NA
  team_db_sim$games_4  <- NA
  team_db_sim$games_2  <- NA
  
  # --- ROUND 1: Round of 64 (32 Games) ---
  for(i in 1:32){
    match_teams <- team_db_sim[which(team_db_sim$games_64 == i),]
    
    if(nrow(match_teams) == 2) {
      prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
      sim_val <- runif(1)
      
      # Determine Winner (1 = Team 1 wins, 0 = Team 2 wins)
      res <- ifelse(sim_val < prob, 1, 0)
      
      # Update Elo Ratings
      new_elo <- elo.calc(res, match_teams$elo_rating[1], match_teams$elo_rating[2], k = k)
      team_db_sim$elo_rating[which(team_db_sim$games_64 == i)] <- as.numeric(new_elo[1, ])
      
      # Assign Winner to Next Round (Round of 32)
      # Logic: Game i feeds into Game ceiling(i/2)
      winner_idx <- which(team_db_sim$games_64 == i)[ifelse(res == 1, 1, 2)]
      team_db_sim$games_32[winner_idx] <- ceiling(i / 2)
    }
  }
  
  # --- ROUND 2: Round of 32 (16 Games) ---
  for(i in 1:16){
    match_teams <- team_db_sim[which(team_db_sim$games_32 == i),]
    
    if(nrow(match_teams) == 2) {
      prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
      sim_val <- runif(1)
      res <- ifelse(sim_val < prob, 1, 0)
      
      new_elo <- elo.calc(res, match_teams$elo_rating[1], match_teams$elo_rating[2], k = k)
      team_db_sim$elo_rating[which(team_db_sim$games_32 == i)] <- as.numeric(new_elo[1, ])
      
      winner_idx <- which(team_db_sim$games_32 == i)[ifelse(res == 1, 1, 2)]
      team_db_sim$games_16[winner_idx] <- ceiling(i / 2)
    }
  }
  
  # --- ROUND 3: Sweet 16 (8 Games) ---
  for(i in 1:8){
    match_teams <- team_db_sim[which(team_db_sim$games_16 == i),]
    
    if(nrow(match_teams) == 2) {
      prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
      sim_val <- runif(1)
      res <- ifelse(sim_val < prob, 1, 0)
      
      new_elo <- elo.calc(res, match_teams$elo_rating[1], match_teams$elo_rating[2], k = k)
      team_db_sim$elo_rating[which(team_db_sim$games_16 == i)] <- as.numeric(new_elo[1, ])
      
      winner_idx <- which(team_db_sim$games_16 == i)[ifelse(res == 1, 1, 2)]
      team_db_sim$games_8[winner_idx] <- ceiling(i / 2)
    }
  }

  # --- ROUND 4: Elite 8 (4 Games) ---
  for(i in 1:4){
    match_teams <- team_db_sim[which(team_db_sim$games_8 == i),]
    
    if(nrow(match_teams) == 2) {
      prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
      sim_val <- runif(1)
      res <- ifelse(sim_val < prob, 1, 0)
      
      new_elo <- elo.calc(res, match_teams$elo_rating[1], match_teams$elo_rating[2], k = k)
      team_db_sim$elo_rating[which(team_db_sim$games_8 == i)] <- as.numeric(new_elo[1, ])
      
      winner_idx <- which(team_db_sim$games_8 == i)[ifelse(res == 1, 1, 2)]
      team_db_sim$games_4[winner_idx] <- ceiling(i / 2)
    }
  }
  
  # --- ROUND 5: Final Four (2 Games) ---
  for(i in 1:2){
    match_teams <- team_db_sim[which(team_db_sim$games_4 == i),]
    
    if(nrow(match_teams) == 2) {
      prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
      sim_val <- runif(1)
      res <- ifelse(sim_val < prob, 1, 0)
      
      new_elo <- elo.calc(res, match_teams$elo_rating[1], match_teams$elo_rating[2], k = k)
      team_db_sim$elo_rating[which(team_db_sim$games_4 == i)] <- as.numeric(new_elo[1, ])
      
      winner_idx <- which(team_db_sim$games_4 == i)[ifelse(res == 1, 1, 2)]
      team_db_sim$games_2[winner_idx] <- 1
    }
  }
  
  # --- ROUND 6: Championship (1 Game) ---
  match_teams <- team_db_sim[which(team_db_sim$games_2 == 1),]
  if(nrow(match_teams) == 2) {
      prob <- elo.prob(match_teams$elo_rating[1], match_teams$elo_rating[2])
      sim_val <- runif(1)
      res <- ifelse(sim_val < prob, 1, 0)
      
      # Determine Champion
      winner_row <- match_teams[ifelse(res == 1, 1, 2), ]
      
      # Store result in list
      results_list[[s]] <- winner_row$team_name
  }
}


results_list_df <- data.frame(champion = unlist(results_list))


print(head(sort(table(results_list_df$champion), decreasing=TRUE), 10))
```

```{r}
#use the original 'tourney_field' to get the full list of team names

champ_counts <- rep(0, nrow(tourney_field))
names(champ_counts) <- tourney_field$team_name


for(i in 1:n_sims){
  
  champion <- results_list[[i]]
  

  if(length(champion) > 0){
    champ_counts[champion] <- champ_counts[champion] + 1
  }
}


champ_prob <- data.frame(
  team = names(champ_counts),
  wins = as.numeric(champ_counts),
  probability = as.numeric(champ_counts) / n_sims
)


champ_prob <- champ_prob[order(champ_prob$wins, decreasing = TRUE),]


head(champ_prob, 10)
```

```{r}

ggplot(head(champ_prob, 10), aes(x = reorder(team, probability), y = probability)) +
  geom_bar(stat = "identity", fill = "#CC5500") +
  coord_flip() +
  labs(
    title = "March Madness Championship Probabilities",
    subtitle = paste("Based on", n_sims, "simulations"),
    x = "Team",
    y = "Probability of Winning"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10)
  )
```

```{r}
save(espn_mbb_pbp, temp_1, temp_2, games_1, games_2, mbb_2026, mbb_2025, df_elo_25, df_elo_26, conferences_26_clean, schedule_26, ACC_Conference, ACC_final_simulated_elos, ACC_remaining_schedule, A10_Conference, A10_final_simulated_elos, A10_remaining_schedule, B1G_Conference, B1G_final_simulated_elos, B1G_remaining_schedule, MAAC_Conference, MAAC_remaining_schedule, merged_mbb_clean, r64_matchups, season_games, projected_standings, final_elos, ranking_data, tourney_field, team_db_sim, results_list_df, champ_counts, champ_prob, file = "mbb_data_4.rda")
```

